Analysis of pudding following Davidson 1970

### 2021-11-04 inst/pudding.Rmd
work

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
library("PlackettLuce")
library("partitions")
options("digits" = 5)
```

@@

```{r}
pudding_table <- pudding  # standard hyper2 terminology
pudding_table
```

@@

`hankin` <- function(ph){
  ##  ph a vector of length n+1; first n entries are strengths, last one
  ##  propensity to tie
  n <- length(ph)-1
  m <- allbinom(n,2)
  out <- matrix(NA,nrow=ncol(m),ncol=3)
  colnames(out) <- c("i","j","tie")
  rownames(out) <- apply(m,2,paste,collapse="")
  th <- ph[n+1]
  for(o in seq_len(ncol(m))){
      i <- m[1,o]
      j <- m[2,o]
      dh <- ph[i] + ph[j] + th    # dh == 'denominator hankin'
      out[o,1] <- ph[i]/dh
      out[o,2] <- ph[j]/dh
      out[o,3] <- th/dh
  }
  return(out)
}

@@

  ##  pd a vector of length n+1; first n entries are strengths, last one
  ##  nu
  n <- length(pd)-1  # how many puddings
  nu <- pd[n+1]  
  m <- allbinom(n,2)
  out <- matrix(NA,nrow=ncol(m),ncol=3)
  colnames(out) <- c("i","j","tie")

  rownames(out) <- apply(m,2,paste,collapse="")

  nu <- pd[n+1]

  for(o in seq_len(ncol(m))){
      i <- m[1,o]
      j <- m[2,o]
      dd <- pd[i] + pd[j] + nu*sqrt(pd[i]*pd[j])  # dd == 'denominator davidson'
      out[o,1] <- pd[i]/dd
      out[o,2] <- pd[j]/dd
      out[o,3] <- nu*sqrt(pd[i]*pd[j])/dd
  }
  return(out)
}

@@

```{r}
n <- pudding_table[,-(1:3),]
h <- hankin(ph)
d <- davidson(pd)
n
h
d
sum(n*log(h))
sum(n*log(d))
sum(n*log(h))-sum(n*log(d))
```

@@

(above, we doctor the dataset so that any comparison involving pudding
1 has a high probability of tying).

Then:

@@

Further, test the null the pudding 1 has zero tying probability:

```{r,thpzero,cache=TRUE}
(tHpzero <- specificp.test(Hp,"tie_3"))
```

@@

```{r,synthmaker}
set.seed(2)
f <- function(i){rmultinom(1,50,prob=h[i,])} # f(i) = sample from row i of h
synth <- pudding_table
synth[,-(1:3)] <- t(sapply(1:15,f))  # random sample
synth[,3] <- 50
rownames(synth) <- rownames(p)
colnames(synth) <- colnames(p)
synth
```

@@

The above strengths are different from (but not too different from)
the value obtaine from `maxp(H)` above.  Now analyse the same dataset
but using the method given in the `PlackettLuce` vignette:

```{r}
PlackettLuce_coef <- function(pudding_table){
i_wins <- data.frame(Winner = pudding_table$i, Loser = pudding_table$j)
j_wins <- data.frame(Winner = pudding_table$j, Loser = pudding_table$i)
  ties <- data.frame(Winner = asplit(pudding_table[c("i", "j")], 1),
                     Loser = rep(NA, 15))
R <- as.rankings(rbind(i_wins, j_wins, ties),
                 input = "orderings")
w <- unlist(pudding_table[c("w_ij", "w_ji", "t_ij")])
mod <- PlackettLuce(R, weights = w, npseudo = 0, maxit = 7)
return(coef(mod,log=FALSE))
}	
```

@@

```{r}
(h_synth <- hankin(ph_synth))
(d_synth <- davidson(pd_synth))
```

@@

```{r}
(n <- synth[,-(1:3),])
(lh_synth <- sum(n*log(h_synth)))
(ld_synth <- sum(n*log(d_synth)))
lh_synth - ld_synth
```

So reified Bradley-Terry is better than Davidson, but (in this case at
least) not by much.

>>>>

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
library("PlackettLuce")
library("partitions")
options("digits" = 5)
showsome <- function(D){
jj <- round(D[c(1:4,nrow(D)),],4)
jj[4,] <- "..."
rownames(jj)[4] <- "..."
noquote(jj)
}
```

@@

```{r}
pudding_table <- pudding  # standard hyper2 terminology
rownames(pudding_table) <- apply(allbinom(6,2),2,paste,collapse="v")
pudding_table
```

@@

`hankin` <- function(ph){
  ##  ph a vector of length n+1; first n entries are strengths, last one
  ##  propensity to tie
  n <- length(ph)-1
  m <- allbinom(n,2)
  out <- matrix(NA,nrow=ncol(m),ncol=3)
  colnames(out) <- c("i","j","tie")
  rownames(out) <- apply(m,2,paste,collapse="v")
  th <- ph[n+1]
  for(o in seq_len(ncol(m))){
      i <- m[1,o]
      j <- m[2,o]
      dh <- ph[i] + ph[j] + th    # dh == 'denominator hankin'
      out[o,1] <- ph[i]/dh
      out[o,2] <- ph[j]/dh
      out[o,3] <- th/dh
  }
  return(out)
}

@@

  ##  pd a vector of length n+1; first n entries are strengths, last one
  ##  nu
  n <- length(pd)-1  # how many puddings
  nu <- pd[n+1]  
  m <- allbinom(n,2)
  out <- matrix(NA,nrow=ncol(m),ncol=3)
  colnames(out) <- c("i","j","tie")
  rownames(out) <- apply(m,2,paste,collapse="v")
  nu <- pd[n+1]
  for(o in seq_len(ncol(m))){
      i <- m[1,o]
      j <- m[2,o]
      dd <- pd[i] + pd[j] + nu*sqrt(pd[i]*pd[j])  # dd == 'denominator davidson'
      out[o,1] <- pd[i]/dd
      out[o,2] <- pd[j]/dd
      out[o,3] <- nu*sqrt(pd[i]*pd[j])/dd
  }
  return(out)
}

@@

```{r}
n <- pudding_table[,-(1:3)]
h <- hankin(ph)
d <- davidson(pd)
showsome(n)
showsome(h)
showsome(d)
sum(n*log(h))
sum(n*log(d))
(suppdiff <- sum(n*log(d))-sum(n*log(h)))
```

Above we see that Davidson's method has almost 3 units of support more
than RBT.  However, we note that Davidson's model has seven degrees of
freedom while RBT has only six (because of the the unit sum
constraint).  Here Wilks's theorem is not applicable because the two
models are not nested but even so the likelihood ratio is a meaningful
statistic.

```{r}
pchisq(suppdiff,df=1,lower.tail=FALSE)
```

@@

(above, we doctor the dataset so that any comparison involving pudding
1 has a high probability of tying).  Then:

@@

Above analysis shows that we can reject the null that all puddings
have the same tying propensity.

@@

```{r,synthmaker}
set.seed(0)
f <- function(i){rmultinom(1,50,prob=h[i,])} # f(i) = sample from row i of h
synth <- pudding_table
synth[,-(1:3)] <- t(sapply(1:15,f))  # random sample
synth[,3] <- 50
rownames(synth) <- rownames(p)
colnames(synth) <- colnames(p)
synth
```

@@

The above strengths are different from (but not too different from)
the value obtained from `maxp(H)` above.  Now analyse the same dataset
but using the method given in the `PlackettLuce` vignette:

```{r}
PlackettLuce_coef <- function(pudding_table){
i_wins <- data.frame(Winner = pudding_table$i, Loser = pudding_table$j)
j_wins <- data.frame(Winner = pudding_table$j, Loser = pudding_table$i)
  ties <- data.frame(Winner = asplit(pudding_table[c("i", "j")], 1),
                     Loser = rep(NA, 15))
R <- as.rankings(rbind(i_wins, j_wins, ties),
                 input = "orderings")
w <- unlist(pudding_table[c("w_ij", "w_ji", "t_ij")])
mod <- PlackettLuce(R, weights = w, npseudo = 0, maxit = 70)
return(coef(mod,log=FALSE))
}	
```

@@

```{r}
showsome(h_synth <- hankin(ph_synth))
showsome(d_synth <- davidson(pd_synth))
```

@@

```{r}
showsome(n <- synth[,-(1:3),])
(lh_synth <- sum(n*log(h_synth)))  # Hankin; RBT
(ld_synth <- sum(n*log(d_synth)))  # Davidson
lh_synth - ld_synth
```

So reified Bradley-Terry is better than Davidson, but (in this case at
least) not by much.  The difference is small because the puddings have
very similar strengths.  We can repeat the analysis but with puddings
of different strengths:

```{r}
dpudstrengths <- c(zipf(6)*0.9,tie=0.1)  # different pudding strengths
names(dpudstrengths) <- c(puddings,"tie")
dpudstrengths
```

Above we see that `pudding_1` is six times stronger than `pudding_6`
in accordance with Zipf's law.  Same analysis as before:

```{r}
showsome(h2 <- hankin(dpudstrengths))
f <- function(i){rmultinom(1,50,prob=h2[i,])} 
synth2 <- pudding_table
synth2[,-(1:3)] <- t(sapply(1:15,f))  # random sample
synth2[,3] <- 50
rownames(synth2) <- rownames(p)
colnames(synth2) <- colnames(p)
showsome(synth2)
```

OK so `synth2` is a different synthetic dataset but this time with
pudding strengths following Zipf.  As before, create a `hyper2`
object and find the evaluate:

```{r,synth2maxp,cache=TRUE}
H_synth2 <- convert(synth2)[[1]] # a hyper2 likelihood function
ph_synth2 <- maxp(H_synth2)
```

```{r}
ph_synth2
```

As a consistency check compare estimate with true value:

```{r}
ph_synth2-dpudstrengths  # difference is small
```

Now Davidson estimate using `synth2`:

```{r,synth2david,cache=TRUE}
(pd_synth2 <- PlackettLuce_coef(synth2))
```

```{r}
h_synth2 <- hankin(ph_synth2)
d_synth2 <- davidson(pd_synth2)
showsome(h_synth2)  # same as before
```

And calculate a support function:

```{r}
n <- synth2[,-(1:3),]  # as before, extract just the counts
(lh_synth2 <- sum(n*log(h_synth2)))
(ld_synth2 <- sum(n*log(d_synth2)))
lh_synth2 - ld_synth2
```

So this time the difference in support is larger, due to the puddings
having different strengths.
